Development of a Python Package for Matching Observational Data

### 2022-05-05 paper/main.tex
rough draft of paper finished

\title{Development of a Python Package for Matching Observational Data}
\author{Jack Potrykus}
\date{\today}

@@

@@

\maketitle{}

@@

@@

Observational studies are of ever-increasing importance as the numerous amount of data collected each year continues to grow.
However, of observational studies are at a major disadvantage compared to blocking-based study design or randomized control trials (RCTs), because the researchers do not have control over the distribution of confounding variables, known or unknown, in the data itself. As such, researchers must take preliminary steps to alter the data in some way in order to minimize the effects of selection bias, which results from heterogeneous distributions present in the observational data.

Within the field of causal inference, the term \emph{matching} is often associated with observational studies with a binary treatment indicator.
Using a similar framework to that of \textcite{iacus_multivariate_2011}, consider $\mat{X}_T \in \reals^{n \times k}$ and $\mat{X}_C \in \reals^{m \times k}$, matrices of $k$ features for $n$ treatment-group observations and $m$ control-group observations, $n \leq m$.
We seek to produce a matching $\set{M}$ such that each observation in $\mat{X}_T$ is paired with one or more nearby observations from $\mat{X}_C$, according to a distance metric and a matching algorithm.

@@

I then introduce a Python package I have developed, \texttt{matching}, which provides an idiomatic and flexible tool for matching observational data. It achieves this by allowing distance measures and matching algorithms to be independently and iteratively applied, so that even incredibly bespoke matching procedures may be implemented succinctly in code. It is also the first Python library for matching observational data which is ``graph-centric'': the key data structure used is a bipartite graph, whose nodes represent observation IDs, and edge weights the distance measure between them. This enables further exploration into new matching procedures and balance diagnostics using graph metrics and algorithms from computer science.

@@

In the context of this paper, distance measures are functions $d: \reals^k \times \reals^k \mapsto \reals^+$ which produce a non-negative distance, or ``cost of matching'', between any vectors $\vec{x}_T, \vec{x}_C$ from the rows of $\mat{X}_T \in \reals^{n \times k}$ and $\mat{X}_C \in \reals^{m \times k}$, to judge the match quality. Distance measures are often combined with some preprocessing function $f(\mat{X})$, usually in the form of a dimension reduction or discretization of the data. Again using a framework akin to that of \textcite{iacus_multivariate_2011}, one may succinctly describe a distance measurement as
\begin{equation}
  \mathcal{D}_d (f(\mat{X}_T), f(\mat{X}_C)), \label{eqn:abstractdistance}
\end{equation}
where $\mathcal{D}_d : \reals^{n \times k} \times \reals^{m \times k} \mapsto \reals^{n \times m}$ is a ``vectorized'' version of $d$, producing a matrix $\mat{D}$ whose $(i, j)$th element is the distance between observation $i$ from $\mat{X}_T$ and observation $j$ from $\mat{X}_C$.

In the simplest case, $d$ is some norm (perhaps Euclidean) calculated on the difference between feature vectors, and $f$ is the identity. With this simple framework in mind, we are now ready to explore extensions of the simple case, and examine their proposed benefits.

@@

Balancing scores
\footnote{Balancing scores come in a variety of forms: as \textcite{rosenbaum_central_1983} note, $b(X) = X$ is the simplest balancing score. However, this paper uses the term ``balancing score'' specifically with regards to 1-dimensional reductions of the data. That is to say, that we will consider balancing scores as a preprocessing function $b: \reals^{(n + m) \times k} \mapsto \reals^{n + m}$.}
\parencite{rosenbaum_central_1983} are a versatile class of functions $b(X)$ satisfying the property
\begin{equation}
  \mat{X} \perp \vec{z} | b(\mat{X}), \label{eqn:defbalancingscore}
\end{equation}
where the matrix $\mat{X} \in \reals^{(n + m) \times k} = \mat{X}_T \cup \mat{X}_C$, and the vector $\vec{z} \in \reals^{n + m}$ contains binary treatment group assignments.
In words, given the balancing score $b(\mat{X})$, the distribution of the feature vector $\mat{X}$ is independent of treatment assignment $\vec{z}$. These methods each set $d$ equal to the $L^1$ norm, or absolute, distance by convention. In the notation of~\eqref{eqn:abstractdistance}, then, balancing scores compose the class of distance measures of the form
\begin{equation}
  \mathcal{D}_{L^1}(b(\mat{X})_T, b(\mat{X})_C).
\end{equation}

TODO DISCUSS STRONG IGNORABILITY, PROPERTIES, WHEN IS ATE ESTIMATE UNBIASED
+ TODO MAKE CLEAR IT IS REPLICATING A RANDOMIZED TRIAL

@@

\paragraph{Joint Use of Balancing Scores.} \textcite{leacy_joint_2014} further explores of calculating \emph{both} propensity and prognostic scores, stratifying each into a $k \times k$ grid, and considering observations within each of the $k^2$ grid squares as a \emph{matched strata}. However, they found that matching on prognostic scores alone performed the best, while matching on propensity scores and Mahalanobis distance (see \S\ref{subsubsec:normdist}) also performed admirably. The $k \times k$ strata performed better than stratification of either score alone, but could not beat matching on either un-discretized score alone. They also found that the percent bias of estimates created using this matching method increased with the non-linearity in the outcomes model.

@@

\paragraph{Criticism.} The most notable criticism of balancing score based distance measurement comes from \textcite{king_why_2019}. King and Nielsen's argument is twofold: first, from an experimental design perspective, they argue that \emph{fully blocked} study design, which other distance measures attempt to emulate (in particular: Mahalanobis distance, Coarsened Exact Matching), dominates \emph{complete randomization study design}, which balancing-score based methods attempt to emulate.
Whereas each study design balances unobserved confounding features on average, fully blocked methods assure balance of observed confounding features, whereas complete randomization only promises balance \emph{on average}.
They thus regard balancing score methods as having a ``lower standard''.
Second, they observe the ``paradox of Propensity Score Matching'': consider the case where $\mat{X} \perp \vec{z}$, and so all predicted scores are the same constant: $\hat{\vec{z}}$.
Then, all potential matches are indecipherable, and the individual matchings may be complete nonsensical. The authors nonetheless concede that propensity scores (and other balancing scores, by extension) do indeed have nice theoretical properties, and may serve as a useful diagnostic; they contend only that they should not be used for matching.

TODO VERIFY

@@

\textcite{iacus_multivariate_2011} introduces a formal definition for a class of distance measures which they call ``monotonic imbalance bounding'', or MIB. These methods are designed to require no assumptions about the distribution of the data, and to emphasize minimizing \emph{in-sample} imbalance, as opposed to \emph{sample} balance. These methods take a vector of tuning parameters $\vec{\uppi}$, such as a vector of calipers. These methods afford the benefit that the number of matches is a function of the tuning parameters alone, and given monotonicity of the maximum distance function $\gamma_{D_{d}}(\vec{\uppi})$, can be each adjusted independently other others: if $\gamma$ is vector-valued, returning a vector in $\reals^k$, the distance between each feature vector can be independently compared in a sequence of inequalities. CEM is an example of an MIB distance measure, but this class is clearly much more flexible, allowing for precise adjustments.

\paragraph{Almost Exact Matching.} Some of the most modern methods are DAME \parencite{liu_interpretable_2019} and FLAME \cite{wang_flame_2021}, which incorporate machine-learning generated weight vectors indicating the relative importance of features to extend exact matching into the field of \emph{almost exact matching}. These two algorithms are unified by their tendency to sometimes ``drop'' uninformative features from the matching process, in order to ensure that the most important features are matched on, and potentially increase the number of potential matches.

TODO FINISH

@@

@@

Once the distance measure has been agreed upon, the researcher must then decide \emph{how} to match. A computer scientist could describe this problem as finding the minimum weight full matching of a \emph{bipartite graph}, whose nodes represent observation IDs, and whose edge weights represent the distance between them. Here, the term ``bipartite'' means that all edges in the graph map $T \leftrightarrow C$; within each disjoint subset, no two nodes share an edge. This paper will explore the optimal solution to this problem in both the $1:1$ and $1:k$ matching scenarios, as well as fast approximations.

@@

\textcite{rosenbaum_optimal_1989} argues for optimal matching according to the Hungarian algorithm, also known as the Kuhn-Munkres or simply Munkres algorithm \parencite{munkres_algorithms_1957}. We will denote a matching a set $\set{M}$ of tuples $(i, j)$, with $i$ representing a treatment-group observation ID, and $j$ a control-group observation ID. The Hungarian algorithm solves the \emph{linear assignment problem},which in this setting may be expressed as
\begin{equation}
  \min_{\set{M}} \sum_{(i, j) \in \set{M}} d(\vec{x}_{T_{i}}, \vec{x}_{C_{j}}), \text{ such that } \forall (i, j), (k, l) \in \set{M}, i = k \iff j = l;
\end{equation}
the ``such that'' condition simply specifies that the matching is one-to-one. As such, optimal matching via the Hungarian algorithm minimizes the total sum of distances between matches in the matched subset.

@@

There are also many graphical diagnostics for assessing balance. QQ-plots, eCDFs, and density estimates for for the unmatched data and the matched subset, perhaps overlaid, effectively visually convey the balance improvement from matching. One may also plot the improvement in standardized mean differences between the two groups, referred to as a Love plot \parencite{greifer_assessing_2022}. One could also plot the bipartite graph to understand where clusters are appearing, or how large groups of ``similar'' observations are, which observations are most ``dissimilar'' from the rest.

@@

After exploring \texttt{MatchIt}'s capabilities, as well as the Python landscape of matching packages for causal inference, I set out to create a package that is:
\begin{itemize}
  \item inspectable. The user should be able to numerically and graphically inspect each step of the matching procedure.
  \item flexible. The package should provide ``plug-and-play'' tools to customize and conduct an arbitrarily complex matching procedure, without the user needing to implement new classes.
  \item familiar. The package should mimic idioms from popular data science packages\footnote{e.g. \texttt{numpy}, \texttt{pandas}, \texttt{sklearn}, \texttt{networkx}}, and all outputs should be easily coercible to a \texttt{pandas.DataFrame} or a \texttt{numpy.ndarray} for further analysis.
  \item ``depth over breadth''. The package should only implement matching procedures, leaving model-building and effect estimation to the user. Narrowing in on the matching procedure alone also helps with the first three goals.
\end{itemize}
In particular, I focused on separating responsibility between how the distance between feature vectors is calculated, and how the matches are then assigned. Being able to easily adjust distance measure and matching procedure independently is especially valuable when assessing the matching quality via diagnostic tools when tuning a hyperparameter, e.g. the maximum allowable distance, or the number of matches $k$ to find for each treatment observation.

@@

\begin{figure}[h!]
  \center{%
    \includegraphics[width=\textwidth]{../docs/mermaid/img/relationships.png}%
  }
  \caption{\label{fig:arch_flow} The relationships between the different modules of \texttt{matching}}.
\end{figure}

@@

\paragraph{Initialization.} 
The \texttt{MatchingGraph} is minimally initialized with an array-like (e.g., a \texttt{pandas.DataFrame}, or a \texttt{numpy.ndarray}) of features \texttt{X} and an array-like of binary treatment assignments \texttt{z}. The user may first wish to preprocess their data (i.e. calculating propensity scores) using \texttt{matching.preprocessing}, but this is not necessary.
\begin{minted}[linenos, fontsize=\small]{python3}
from matching.graph import MatchingGraph
from matching.preprocessing import propensity_score

# Possibly some preprocessing... (not shown)
# X is an array of propensity scores, z is an array of booleans
mg = MatchingGraph(X, z)
\end{minted}
\paragraph{Setting Edges.} The user should then make a call to the \texttt{set\_edges} method to calculate the distance between each observation the two groups, and set the graph nodes and edge weights. Here, the user supplies a \texttt{\_Distance} measure from \texttt{matching.distance}. The user may supply a \texttt{min\_distance} and/or a \texttt{max\_distance} for an edge to be allowed; in graph terms, this can be considered the edges' \emph{capacity}.
\begin{minted}[linenos, fontsize=\small]{python3}
import numpy as np
from matching.distance import L1Norm

# Calculating the caliper width. Recall that X is an array of scores
caliper = 0.05
caliper_width = caliper * np.std(X)
mg.set_edges(distance=L1Norm(max_distance=caliper_width))
\end{minted}
\paragraph{Iterative Filtering.} It is possible to implement iterative distance measures as well. Consider \texttt{X} as a dataframe with columns \texttt{"score"}, containing propensity scores, and \texttt{"is\_nice"}, a boolean series. We would like to match on propensity scores within \texttt{caliper\_width}, but also cannot assign any nice people to any naughty people. We can make use of the \texttt{filter\_edges} method and the \texttt{include} keyword argument to accomplish this.
\begin{minted}[linenos, fontsize=\small]{python3}
from matching.distance import Exact, L1Norm
from matching.graph import MatchingGraph
from matching.preprocessing import propensity_score

@@

There are also methods to \texttt{filter\_nodes} by label or order, as well as to \texttt{filter\_subgraphs}, which filters the disjoint subgraphs of the graph by a variety of metrics\footnote{\texttt{filter\_subgraphs} is particularly relevant for strata-based matching methods like CEM, which often filter strata based on the total number of observations (from each group), the ratio of treatment to control}; they will not be discussed here, but readers are referred to \texttt{matching} documentation hosted on the GitHub\footnote{\url{https://github.com/jackpotrykus/propensity-score-matching-thesis}}.

\paragraph{Matching.} Once the user has set the edges of the graph and completed their filtering, they may wish to conduct a $1:k$ matching between treatment and control, for some integer $k \geq 1$\footnote{This is not always necessary, particularly in the case of CEM}.
The user does this via a call to the \texttt{match} method. The user can control $k$ with \texttt{n\_match}, the maximum number of matches to find for each observation. Some treatment observations will not be able to be matched with this many control observations; they can be automatically pruned with the \texttt{min\_match} parameter. The parameter \texttt{replace} is a boolean, indicating whether multiple treatment observations may match with the same control observation, which is \texttt{False} by default. Finally, the user has a choice of greedy (keyword argument: \texttt{"greedy"} or \texttt{"fast"}) or optimal matches (keyword argument: \texttt{"optimal"}, \texttt{"hungarian"}, \texttt{"kuhn"}, or \texttt{"munkres"}).
\begin{minted}[linenos, fontsize=\small]{python3}
# Now we should match the graph
# By default: n_match=1, min_match=1, replace=False
mg.match(n_match=2, min_match=2, method="optimal")
\end{minted}

\paragraph{Balance Assessment.} The user can then compare the balance of the \texttt{input\_data} and the \texttt{match\_data}\footnote{In the case of Exact matching edges, the \texttt{match\_data} should consist of all subgraphs of order at least 2}. Note that \texttt{match\_data} is a subset of \texttt{input\_data}, plus a new column, \texttt{"match\_group"}.
\begin{minted}[linenos, fontsize=\small]{python3}
# Can get dataframes of input_data and match_data
input_df = mg.input_data.frame
match_df = mg.match_data.frame

@@

@@

\paragraph{\texttt{dame-flame}.} Developed by Duke's \emph{Almost Matching Exactly Lab} \parencite{gupta_dame-flame_2021}, \texttt{dame-flame} implements its two titular matching procedures: DAME \parencite{liu_interpretable_2019} and FLAME \parencite{wang_flame_2021}. As discussed in \S\ref{subsubsec:exactmatching}, these methods extend CEM using machine learning, calculating a weight vector \vec{w} indicating the relative importance of matching on each feature. These algorithms each run fast, and are easier to inspect than \texttt{DoWhy}. However, since these each seek to replicate CEM, and since these are the \emph{only} matching methods provided by \texttt{dame-flame}, this means that \texttt{dame-flame} can only be used with discrete datasets; for users unwilling to coarsen/discretize their data, or users seeking a matching procedure other than DAME and FLAME, \texttt{dame-flame} is not an option. \texttt{matching} does not currently support these algorithms, but they are ``in-the-works''; it can certainly be implemented as a new \texttt{matching.distance.\_Distance} subclass.

\section{Experiments}

>>>>

@@

\title{Development of a Python Package for Matching Observational Data}
\subtitle{Advisors: Jingshu Wang, Sanjay Krishnan}
\author{Jack Potrykus}
\date{\today}

@@

% Insert some text making clear that this is an MS thesis
\begin{center}{\large \bfseries Master’s Paper for the University of Chicago Department of Statistics}\end{center}

% Relax `\maketitle` to allow text on same page. https://tex.stackexchange.com/questions/86249/maketitle-text-before-title
{\let\newpage\relax\maketitle}

@@

% Insert a spot for the advisor's signature
\vfill{\null}
Approved by: \hrulefill~Date: \hrulefill

@@

Observational studies are of ever-increasing importance as the numerous amount of data collected each year continues to grow.
However, of observational studies are at a major disadvantage compared to blocking-based study design or randomized control trials, because the researchers do not have control over the distribution of confounding variables, known or unknown, in the data itself. As such, researchers must take preliminary steps to alter the data in some way in order to minimize the effects of selection bias, which results from heterogeneous distributions present in the observational data.

Within the field of causal inference, the term \emph{matching} is often associated with observational studies with a binary treatment indicator.
Using a similar framework to that of \textcite{iacus_multivariate_2011}, consider $\mat{X}_T \in \reals^{n \times p}$ and $\mat{X}_C \in \reals^{m \times p}$, matrices of $p$ features for $n$ treatment-group observations and $m$ control-group observations, $n \leq m$.
We seek to produce a matching $\set{M}$ such that each observation in $\mat{X}_T$ is paired with one or more nearby observations from $\mat{X}_C$, according to a distance metric and a matching algorithm.

@@

I then introduce a Python package I have developed, \texttt{matching}, which provides an idiomatic and flexible tool for matching observational data. It achieves this by allowing distance measures and matching algorithms to be independently and iteratively applied, so that even incredibly bespoke matching procedures may be implemented succinctly in code. It is also the first Python library for matching observational data which is ``graph-centric'': the key data structure used is a bipartite graph, whose nodes represent observations, and edge weights the distance measure between them. This enables further exploration into new matching procedures and balance diagnostics using graph metrics and algorithms from computer science.

@@

In the context of this paper, distance measures are functions $d: \reals^p \times \reals^p \mapsto \reals^+$ which produce a non-negative distance, or ``cost of matching'', between any vectors $\vec{x}_T, \vec{x}_C$ from the rows of $\mat{X}_T \in \reals^{n \times p}$ and $\mat{X}_C \in \reals^{m \times p}$, to judge the match quality. Distance measures are often combined with some preprocessing function $f(\mat{X})$, usually in the form of a dimension reduction or discretization of the data. Again using a framework akin to that of \textcite{iacus_multivariate_2011}, one may succinctly describe a distance measurement as
\begin{equation}
  \mathcal{D}_d (f(\mat{X}_T), f(\mat{X}_C)), \label{eqn:abstractdistance}
\end{equation}
where $\mathcal{D}_d : \reals^{n \times p} \times \reals^{m \times p} \mapsto \reals^{n \times m}$ is a ``vectorized'' version of $d$, producing a matrix $\mat{D}$ whose $(i, j)$th element is the distance between observation $i$ from $\mat{X}_T$ and observation $j$ from $\mat{X}_C$.

@@

Balancing scores
\footnote{Balancing scores come in a variety of forms: as \textcite{rosenbaum_central_1983} note, $b(X) = X$ is the simplest balancing score. However, this paper uses the term ``balancing score'' specifically with regards to 1-dimensional reductions of the data. That is to say, that we will consider balancing scores as a preprocessing function $b: \reals^{(n + m) \times p} \mapsto \reals^{n + m}$.}
\parencite{rosenbaum_central_1983} are a versatile class of functions $b(X)$ satisfying the property
\begin{equation}
  \mat{X} \perp \vec{z} | b(\mat{X}), \label{eqn:defbalancingscore}
\end{equation}
where the matrix $\mat{X} \in \reals^{(n + m) \times p} = \mat{X}_T \cup \mat{X}_C$, and the vector $\vec{z} \in \reals^{n + m}$ contains binary treatment group assignments.
In words, given the balancing scores $b(\mat{X})$, the distribution of the feature matrix $\mat{X}$ is independent of treatment assignment $\vec{z}$. These methods each set $d$ equal to the $L^1$ norm, or absolute distance, by convention. In the notation of~\eqref{eqn:abstractdistance}, then, balancing scores compose the class of distance measures of the form
\begin{equation}
  \mathcal{D}_{L^1}(b(\mat{X})_T, b(\mat{X})_C).
\end{equation}

Balancing scores afford observational data some of the desirable properties of randomized trials. Namely, in a randomized trial, $\mat{X}$ should contain all features used in determining treatment assignment, affording conditional independence of outcomes from treatment:
\begin{equation}
  \vec{y}_0, \vec{y}_1 \perp \vec{z} | \mat{X}, \label{eqn:defstrongignorable}
\end{equation}
where $\vec{y}_0$ and $\vec{y}_1$ are vectors of outcomes, for which the $i$th element is the outcome of observation $i$ when assigned to group 0 (control) or 1 (treatment); note that in practice we almost always only observe $y_{0,i}$ \emph{or} $y_{1,i}$ for each $i$. If the statement in~\eqref{eqn:defstrongignorable} holds, then the treatment assignment is \emph{strongly ignorable}. \textcite{rosenbaum_central_1983} then prove that if~\eqref{eqn:defstrongignorable} holds (that is, treatment $\vec{z}$ is strongly ignorable given features $\mat{X}$), then treatment assignment is \emph{also} strongly ignorable given any balancing score $b(X)$! Conditioning on balancing scores, we gain independence of treatment from features, as well as the strong ignorability of treatment; these useful properties form the key motivation for the use of balancing scores in causal inference.

@@

\paragraph{Joint Use of Balancing Scores.} \textcite{leacy_joint_2014} further explores of calculating \emph{both} propensity and prognostic scores, stratifying each into a $p \times p$ grid, and considering observations within each of the $p^2$ grid squares as a \emph{matched strata}. However, they found that matching on prognostic scores alone performed the best, while matching on propensity scores and Mahalanobis distance (see \S\ref{subsubsec:normdist}) also performed admirably. The $p \times p$ strata performed better than stratification of either score alone, but could not beat matching on either un-discretized score alone. They also found that the percent bias of estimates created using this matching method increased with the non-linearity in the outcomes model.

@@

\paragraph{Criticism.} The most notable criticism of balancing score based distance measurement comes from \textcite{king_why_2019}. King and Nielsen's argument is twofold: first, from an experimental design perspective, they argue that \emph{fully blocked} study design, which other distance measures attempt to emulate (in particular: Mahalanobis distance, Coarsened Exact Matching), dominates \emph{complete randomization study design}, which balancing-score based methods attempt to emulate.
Whereas each study design balances unobserved confounding features on average, fully blocked methods assure balance of observed confounding features, whereas complete randomization only promises balance \emph{on average}.
They thus regard balancing score methods as having a ``lower standard''.
Second, they observe the ``Paradox of Propensity Score Matching'': consider the case where $\mat{X} \perp \vec{z}$, and so all predicted scores are the same constant: $\hat{\vec{z}}$.
Then, all potential matches are indecipherable, and the individual matchings may be complete nonsensical, and in fact may \emph{increase} the imbalance of features within the matched subset. The authors nonetheless concede that propensity scores (and other balancing scores, by extension) do indeed have nice theoretical properties, and may serve as a useful diagnostic; they contend only that they should not be used for matching.

@@

\textcite{iacus_multivariate_2011} introduces a formal definition for a class of distance measures which they call ``monotonic imbalance bounding'', or MIB. These methods are designed to require no assumptions about the distribution of the data, and to emphasize minimizing \emph{in-sample} imbalance, as opposed to \emph{sample} balance. These methods take a vector of tuning parameters $\vec{\uppi}$, such as a vector of calipers. These methods afford the benefit that the number of matches is a function of the tuning parameters alone, and given monotonicity of the maximum distance function $\gamma_{\mathcal{D}_{d}}(\vec{\uppi})$, can be each adjusted independently other others: if $\gamma$ is vector-valued, the element-wise differences of feature vectors can be independently compared in a system of inequalities. CEM is an example of an MIB distance measure, but this class is clearly much more flexible, allowing for precise adjustments.

\paragraph{Almost Exact Matching.} Some of the most modern methods are DAME \parencite{liu_interpretable_2019} and FLAME \cite{wang_flame_2021}, which incorporate machine-learning generated weight vectors indicating the relative importance of features to extend exact matching into the field of \emph{almost exact matching}. These two algorithms are unified by their tendency to sometimes ``drop'' uninformative features from the matching process, in order to ensure that the most important features are matched on, and potentially increase the number of potential matches. The chief difference can be summarized as follows: DAME choosing a subset of features to match on by minimizing predictive error, whereas FLAME maximizes a match quality statistic that the researcher can tune via a hyperparameter; this statistic is a function of predictive error and a \emph{balance factor}, which represents the relative size of the strata of observations matched on this subset on features \parencite{gupta_dame-flame_2021}.

@@

\subsubsection{Iterative Distance Calculations}

Often, researchers will employ an ``iterative'' distance measure. They may, for example, first calculate propensity scores $\widehat{\vec{z}}$, filter potential matches using a caliper $c$, and then match on Mahalanobis distance \emph{within} clusters; or, in the absence of a caliper, within strata of $\widehat{\vec{z}}$. This idea was explored by \textcite{rosenbaum_constructing_1985}, who found it to perform better than matching on either propensity score or Mahalanobis distance alone.

\textcite{baltar_mahalanobis_2014} explored the efficacy of this iterative method in practice, relative to propensity score and Mahalanobis distance alone. In the paper, they compare seek to estimate the effect of Brazilian towns' social agenda on public health. They found both the propensity score alone and iterative method to perform admirably, whereas Mahalanobis distance performed poorly; they do not take this as an indictment of Mahalanobis distance matching, and pose for further research to identify conditions under which Mahalnobis distance may perform \emph{better}. The iterative method achieved the greatest reduction in bias for all features, except the propensity score, for which propensity score distance alone (somewhat obviously) saw the greatest bias reduction. However, as \textcite{austin_optimal_2011} and \textcite{greifer_assessing_2022} note, balance of the scores does not imply balance of the features, and balancing score balance metrics should only be used as a supplementary diagnostic; the focus of balance assessment should be balance of the features.

@@

Once the distance measure has been agreed upon, the researcher must then decide \emph{how} to match. A computer scientist could describe this problem as finding the minimum weight full matching of a \emph{bipartite graph}, whose nodes represent observations, and whose edge weights represent the distance between them. Here, the term ``bipartite'' means that all edges in the graph map $T \leftrightarrow C$; within each disjoint subset, no two nodes share an edge. This paper will explore the optimal solution to this problem in both the $1:1$ and $1:k$ matching scenarios, as well as fast approximations.

@@

\textcite{rosenbaum_optimal_1989} argues for optimal matching according to the Hungarian algorithm, also known as the Kuhn-Munkres or simply Munkres algorithm \parencite{munkres_algorithms_1957}. We will denote a matching a set $\set{M}$ of tuples $(i, j)$, with $i$ representing a treatment-group observations, and $j$ a control-group observations. The Hungarian algorithm solves the \emph{linear assignment problem}, which in this setting may be expressed as
\begin{equation}
  \min_{\set{M}} \sum_{(i, j) \in \set{M}} d(\vec{x}_{T_{i}}, \vec{x}_{C_{j}}), \text{ such that } \forall (i, j), (k, l) \in \set{M}, i = k \iff j = l;
\end{equation}
the ``such that'' condition simply specifies that the matching is one-to-one. As such, optimal matching via the Hungarian algorithm minimizes the total sum of distances between matches in the matched subset.

@@

There are also many graphical diagnostics for assessing balance. QQ-plots, eCDFs, and density estimates for for the unmatched data and the matched subset, perhaps overlaid, effectively visually convey the balance improvement from matching. One may also plot the improvement in standardized mean differences between the two groups, referred to as a Love plot \parencite{greifer_assessing_2022}. One could also plot the bipartite graph to understand where clusters are appearing, or how large groups of ``similar'' observations are, or which observations are most ``dissimilar'' from the rest.

@@

After exploring \texttt{MatchIt}'s capabilities, as well as the Python landscape of matching packages for causal inference, I set out to create a package that is:
\begin{itemize}
  \item inspectable. The user should be able to numerically and graphically inspect each step of the matching procedure.
  \item flexible. The package should provide ``plug-and-play'' tools to customize and conduct an arbitrarily complex matching procedure, without the user needing to implement new classes.
  \item familiar. The package should mimic idioms from popular data science packages\footnote{e.g. \texttt{numpy}, \texttt{pandas}, \texttt{sklearn}, \texttt{networkx}}, and all outputs should be easily coercible to a \texttt{pandas.DataFrame} or a \texttt{numpy.ndarray} for further analysis.
  \item ``depth over breadth''. The package should only implement matching procedures, leaving model-building and effect estimation to the user; narrowing in on the matching procedure alone also helps to achieve the first three goals.
\end{itemize}
In particular, I focused on separating responsibility between how the distance between feature vectors is calculated, and how the matches are then assigned. Being able to easily adjust distance measure and matching procedure independently is especially valuable when assessing the matching quality via diagnostic tools when tuning a hyperparameter, e.g. the maximum allowable distance, or the number of matches $k$ to find for each treatment observation.

@@

\begin{figure}[h!]
  \center{%
    \includegraphics[width=0.8\textwidth]{../docs/mermaid/img/relationships.png}%
  }
  \caption{\label{fig:arch_flow} The relationships between the different modules of \texttt{matching}}.
\end{figure}

@@

\paragraph{Initialization.} 
The \texttt{MatchingGraph} is minimally initialized with an array-like (e.g., a \texttt{pandas.DataFrame}, or a \texttt{numpy.ndarray}) of features \texttt{X} and an array-like of binary treatment assignments \texttt{z}. The user may first wish to preprocess their data (e.g. calculating propensity scores) using \texttt{matching.preprocessing}, but this is not necessary.
\begin{minted}[linenos, fontsize=\small]{python3}
from matching.graph import MatchingGraph
from matching.preprocessing import propensity_score

# Possibly some preprocessing... (not shown)
# X is an array of propensity scores, z is an array of booleans
mg = MatchingGraph(X, z)
\end{minted}
\paragraph{Setting Edges.} The user should then make a call to the \texttt{set\_edges} method to calculate the distance between each observation the two groups, and set the graph nodes and edge weights. Here, the user supplies a \texttt{\_Distance} measure from \texttt{matching.distance}. The user may supply a \texttt{min\_distance} and/or a \texttt{max\_distance} for an edge to be allowed; in graph terms, this can be considered the edges' \emph{capacity}. For the \texttt{PropensityScore} distance metric, \texttt{max\_distance} is instead set by specifying a \texttt{caliper}; the \texttt{max\_distance} is later inferred, once the scores are calculated.
\begin{minted}[linenos, fontsize=\small]{python3}
import numpy as np
from matching.distance import PropensityScore

# Will calculate propensity score distance automatically
mg.set_edges(distance=PropensityScore(caliper=0.05))
\end{minted}
\paragraph{Iterative Filtering.} It is possible to implement iterative distance measures as well. Consider \texttt{X} as a dataframe with columns \texttt{"score"}, containing propensity scores, and \texttt{"is\_nice"}, a boolean series. We would like to match on propensity scores within \texttt{caliper\_width}, but also cannot assign any nice people to any naughty people. We can make use of the \texttt{filter\_edges} method and the \texttt{include} keyword argument to accomplish this.
\begin{minted}[linenos, fontsize=\small]{python3}
from matching.distance import Exact, L1Norm
from matching.graph import MatchingGraph
from matching.preprocessing import propensity_score

@@

There are also methods to \texttt{filter\_nodes} by label or order, as well as to \texttt{filter\_subgraphs}, which filters the disjoint subgraphs of the graph by a variety of metrics\footnote{\texttt{filter\_subgraphs} is particularly relevant for strata-based matching methods like CEM, which often filter strata based on the total number of observations (from each group), the ratio of treatment to control.}; they will not be discussed here, but readers are referred to \texttt{matching} documentation hosted on the GitHub\footnote{\url{https://github.com/jackpotrykus/propensity-score-matching-thesis}}.

\paragraph{Matching.} Once the user has set the edges of the graph and completed their filtering, they may wish to conduct a $1:k$ matching between treatment and control, for some integer $k \geq 1$\footnote{This is not always necessary, particularly in the case of CEM.}
The user does this via a call to the \texttt{match} method. The user can control $k$ with \texttt{n\_match}, the maximum number of matches to find for each observation. Some treatment observations will not be able to be matched with this many control observations; they can be automatically pruned with the \texttt{min\_match} parameter. The parameter \texttt{replace} is a boolean, indicating whether multiple treatment observations may match with the same control observation, which is \texttt{False} by default. Finally, the user has a choice of greedy (keyword argument: \texttt{"greedy"} or \texttt{"fast"}) or optimal matches (keyword argument: \texttt{"optimal"}, \texttt{"hungarian"}, \texttt{"kuhn"}, or \texttt{"munkres"}).
\begin{minted}[linenos, fontsize=\small]{python3}
# Continue from the previous excerpt. Now we want to conduct 1:k matching
# NOTE: By default: n_match=1, min_match=1, replace=False
mg.match(n_match=2, min_match=2, method="optimal")
\end{minted}

\paragraph{Balance Assessment.} The user can then compare the balance of the \texttt{input\_data} and the \texttt{match\_data}\footnote{In the case of Exact matching edges, the \texttt{match\_data} should consist of all subgraphs of order at least 2.}. Note that \texttt{match\_data} is a subset of \texttt{input\_data}, plus a new column, \texttt{"match\_group"}.
\begin{minted}[linenos, fontsize=\small]{python3}
# Can get dataframes of input_data and match_data
input_df = mg.input_data.frame
match_df = mg.match_data.frame

@@

\paragraph{Drawing the Graph.} After edges have been set using \texttt{set\_edges}, users may draw the graph at any point in the matching process. \texttt{networkx} provides a myriad of functions to draw graphs, but the \texttt{MatchingGraph} class has two convenient graph-drawing methods built-in. These are the methods \texttt{draw} and \texttt{draw\_bipartite}. The main difference between these two is that \texttt{draw} will draw the graph such that edge length is proportional to the distance between observations, whereas \texttt{draw\_bipartite} draws the graph in the \emph{bipartite layout}, with all treatment group nodes on one side, and all control group nodes on the other. Example output of each of these is displayed in Figure~\ref{fig:draw_examples}.
\begin{figure}[h!]
    \centering
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[width=0.95\textwidth]{./img/draw_examples/draw.png}
        \caption{Graph produced by \texttt{mg.draw()}. Edge length corresponds to the distance between observations.}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[width=0.95\textwidth]{./img/draw_examples/draw_bipartite.png}
        \caption{Graph produced by \texttt{mg.draw\_bipartite()}. The bipartite structure of the graph is clear, but edge lengths are arbitrary.}
    \end{subfigure}
    \caption{Two visualizations of the \emph{same} graph. This is a graph that resulted from $1:3$ optimal matching. Each graph automatically applies the color-coding of treatment nodes in {\color{red}red} and control nodes in {\color{blue} blue}. Observation IDs/labels may be displayed in either plot by passing \texttt{with\_labels=True}.}
    \label{fig:draw_examples}
\end{figure}

@@

\paragraph{\texttt{dame-flame}.} Developed by Duke's \emph{Almost Matching Exactly Lab} \parencite{gupta_dame-flame_2021}, \texttt{dame-flame} implements its two titular matching procedures: DAME \parencite{liu_interpretable_2019} and FLAME \parencite{wang_flame_2021}. As discussed in \S\ref{subsubsec:exactmatching}, these methods extend CEM using machine learning, calculating a weight vector \vec{w} indicating the relative importance of matching on each feature. These algorithms each run fast, and are easier to inspect than \texttt{DoWhy}. However, since these each seek to replicate CEM, and since these are the \emph{only} matching methods provided by \texttt{dame-flame}, this means that \texttt{dame-flame} can only be used with discrete datasets; for users unwilling to coarsen/discretize their data, or users seeking a matching procedure other than DAME and FLAME, \texttt{dame-flame} is not an option. \texttt{matching} does not currently support these algorithms, but they are ``in-the-works''; it can certainly be implemented as a new \texttt{matching.distance.\_Distance} subclass, and their impressive performance merits their inclusion.

\section{Experiments}
\label{sec:experiments}
